No proof nets for \capsabbrev{mll} with units\\[5pt]\Large Proof equivalence in \capsabbrev{mll} is \capsabbrev{pspace}-complete

### 2014-04-03 main/LICS.tex
added figure, minor edits

Later, a striking technical innovation of linear logic was the introduction of proof nets \cite{Girard-1987,Danos-Regnier-1989}, which define a canonical form for proofs in the unitless fragment of multiplicative linear logic -- two proofs are equivalent if and only if they have the same proof net -- so proof nets offer a simple decision procedure for proofs in this fragment. This naturally raises the question whether proof nets can be extended to work in the presence of units. The work in this direction begins with \cite{Trimble-1994,Blute-Cockett-Seely-Trimble-1996} via \color{red}\cite{Lamarche-Strassburger-2006}\color{black} and perhaps culminates in \cite{HughesFreeStar,HughesMLLProofNets}; but these proof nets are not canonical and must be identified up to a rewiring equivalence.

@@

The permutations of sequent proofs are exactly the identifications imposed by the categorical semantics of \capsabbrev{mll}, star-autonomous categories \cite{Barr-1991} (and semi-star-autonomous categories \cite{Houston-2008} for \capsabbrev{mll} without units).
%
Proof equivalence for \capsabbrev{mll} is therefore equivalent to the \emph{word problem} for star-autonomous categories: the problem whether two representations denote the same morphism in any star-autonomous category.

@@

\aftersubsection
We will use a concise diagrammatic notation for sequents and proof nets.
%
The units $1$ and $\bot$ are represented by a circle $\circ$ and a disc $\bullet$ respectively;
%
formulae related by a tensor will be connected by edges; 
and
formulae related by a par will be juxtaposed, and collected in a box when a par-formula is an immediate subformula of a tensor-formula.
%
For example, the following denote the same sequent:
\[
	\bot\tn\bot,\bot\tn\bot,\1,(\1\parr\1\parr\1)\tn\bot
\]
\[
\begin{tikzpicture}[x=5mm,y=-5mm,octo]
	\draw (0,0) node[bullet] (a) {} -- (1,0) node[bullet] (b) {} (2,0) node[bullet] (c) {} -- (3,0) node[bullet] (d) {};
	\node[circ] (1) at (4,0) {};
	\node[circ] (2) at (5.25,0) {}; \node[circ] (3) at (6.25,0) {}; \node[circ] (4) at (7.25,0) {};
	\draw[rounded corners] (4.75,-1) rectangle (7.75,1);
	\draw (7.75,0) -- (8.75,0) node[bullet] (e) {};
\end{tikzpicture}
\]
The links of a proof net are added to the sequent as coloured arrows.
%
The following example is a proof net for the above sequent.
\[
\begin{tikzpicture}[x=5mm,y=-5mm,octo]
	\node[circ] (1) at (6,0) {};
	\draw (0,0) node[bullet] (a) {} -- (1,0) node[bullet] (b) {} (3,0) node[bullet] (c) {} -- (4,0) node[bullet] (d) {};
	\node[circ] (2) at (1,3) {}; \node[circ] (3) at (2,3) {}; \node[circ] (4) at (3,3) {};
	\draw[rounded corners] (0,2) rectangle (4,4);
	\draw (4,3) -- (5,3) node[bullet] (e) {};
	\begin{scope}[thepink,->]
			\draw[bend right=10] (a) to (2.center);
			\draw[bend right=10] (b) to (3.center);
			\draw[bend left=10]  (c) to (3.center);
			\draw[bend left=10]  (d) to (4.center);
			\draw[bend right=10] (e) to (1.center);
	\end{scope}
\end{tikzpicture}
\]

@@

In a proof net, the \emph{kingdom} and the \emph{empire} of a vertex $v$ are respectively the smallest and largest subnet that have $v$ as a port.
%
In working with the rewiring relation, the notion of empire can be particularly useful.
%
\color{red} 
We will denote the empire of $v$ in $\links$ by $\emp v$.
\color{black}

\begin{proposition}[{\cite[Proposition 2.b]{Bellin-vandeWiele-1995}}]
\label{prop:empire propagation}
The empire $\emp v$ is determined by propagation from $v$: 1) through links; 2) up towards subformulae; 3) into a tensor if one of its subformulae is in $\emp v$; 4) into a par if all its subformulae are in $\emp v$; 5) except never into a par or tensor if $v$ is a subformula.
\end{proposition}

@@

Let a \emph{basic} sequent be one of formulae constructed only over $\1$, $\bot$, and $\tn$.
%
After removing $\bot$-formulae and replacing subformulae $\1\tn A$ with $A$, basic sequents consist of formulae of the form $1$ or $\bigtn(\bot^n)$ with $n\geq2$.

@@

\aftersection
%
\emph{Non-deterministic constraint logic} \cite{Hearn-Demaine-2005,Demaine-Hearn-2008,GamesPuzzlesAndComputation} is a simple graph-rewriting formalism, used here as a convenient tool for \capsabbrev{pspace}-hardness reduction.
%
A \emph{constraint graph} is a graph with weighted edges and an \emph{inflow constraint} -- a natural number that may be taken to be always $2$ -- on each vertex. A \emph{configuration} is an assignment of directions to the edges of the underlying undirected constraint graph. A rewrite step consists of the reversal of a single edge in a configuration:
rewrite steps must preserve the condition that the total weight of the incoming edges at each vertex is at least its inflow constraint.

@@

\begin{definition}
\label{def:standard linkings}
For a weight element $W=W_n(i,j)=A\parr B\parr C$,
\begin{enumerate}
	\item
for a constraint element $C=C_n(i)=X\tn Y$, the \emph{standard linking} for the sequent $W,C$ links the first $\bot$ in $A$ to the first $\1$ in $X$, the first $\bot$ in $B$ and $C$ each to the first $\1$ in $Y$, and each remaining $\bot$ in $A,B,C$ to a remaining $\1$ in $X,Y$ in their order of occurrence;
	\item
for $C=C_n(j)=X\tn Y$, the \emph{standard linking} for $W,C$ is defined as above, except the first $\bot$ in $B$ links to the first $\1$ in $X$;
	\item
the \emph{standard linking} for the sequent $W,\1^{3n+4}$ links the first $\bot$ in $A$, $B$, and $C$ to the first $\1$, and each remaining $\bot$ to a remaining $\1$ in order of occurrence.
\end{enumerate}
\end{definition}

>>>>

Later, a striking technical innovation of linear logic was the introduction of proof nets \cite{Girard-1987,Danos-Regnier-1989}, which define a canonical form for proofs in the unitless fragment of multiplicative linear logic -- two proofs are equivalent if and only if they have the same proof net -- so proof nets offer a simple decision procedure for proofs in this fragment. This naturally raises the question whether proof nets can be extended to work in the presence of units. The work in this direction begins with \cite{Trimble-1994,Blute-Cockett-Seely-Trimble-1996} via \color{red}\cite{Lamarche-Strassburger-2006} \color{black} and perhaps culminates in \cite{HughesFreeStar,HughesMLLProofNets}; but these proof nets are not canonical and must be identified up to a rewiring equivalence.

@@

The permutations of sequent proofs are exactly the identifications imposed by the categorical semantics of \capsabbrev{mll}, star-autonomous categories \cite{Barr-1991} (and semi-star-autonomous categories \cite{Houston-2008,Heijltjes-Strassburger} for \capsabbrev{mll} without units).
%
Proof equivalence for \capsabbrev{mll} is therefore equivalent to the \emph{word problem} for star-autonomous categories: the problem whether two representations denote the same morphism in any star-autonomous category.

@@

\aftersubsection
We will use a concise diagrammatic notation for sequents and proof nets.
%
The units $1$ and $\bot$ are represented by a circle $\circ$ and a disc $\bullet$ respectively;
%
formulae related by a tensor will be connected by edges; 
and
formulae related by a par will be juxtaposed, and collected in a box when a par-formula is an immediate subformula of a tensor-formula \color{red}(or for illustrative purposes)\color{black}.
%
For example, the following denote the same sequent:
\[
	\bot\tn\bot,\bot\tn\bot,\1,(\1\parr\1\parr\1)\tn\bot
\]
\[
\begin{tikzpicture}[x=5mm,y=-5mm,octo]
	\draw (0,0) node[bullet] (a) {} -- (1,0) node[bullet] (b) {} (2,0) node[bullet] (c) {} -- (3,0) node[bullet] (d) {};
	\node[circ] (1) at (4,0) {};
	\node[circ] (2) at (5.25,0) {}; \node[circ] (3) at (6.25,0) {}; \node[circ] (4) at (7.25,0) {};
	\draw[rounded corners] (4.75,-1) rectangle (7.75,1);
	\draw (7.75,0) -- (8.75,0) node[bullet] (e) {};
\end{tikzpicture}
\]
The links of a proof net are added to the sequent as coloured arrows.
%
The following example is a proof net for the above sequent.
\[
\begin{tikzpicture}[x=5mm,y=-5mm,octo]
	\node[circ] (1) at (6,0) {};
	\draw (0,0) node[bullet] (a) {} -- (1,0) node[bullet] (b) {} (3,0) node[bullet] (c) {} -- (4,0) node[bullet] (d) {};
	\node[circ] (2) at (1,3) {}; \node[circ] (3) at (2,3) {}; \node[circ] (4) at (3,3) {};
	\draw[rounded corners] (0,2) rectangle (4,4);
	\draw (4,3) -- (5,3) node[bullet] (e) {};
	\begin{scope}[thepink,->]
			\draw[bend right=10] (a) to (2.center);
			\draw[bend right=10] (b) to (3.center);
			\draw[bend left=10]  (c) to (3.center);
			\draw[bend left=10]  (d) to (4.center);
			\draw[bend right=10] (e) to (1.center);
	\end{scope}
\end{tikzpicture}
\]

@@

In a proof net, the \emph{kingdom} and the \emph{empire} of a vertex $v$ are respectively the smallest and largest subnet that have $v$ as a port.
%
In working with the rewiring relation, the notion of empire can be particularly useful.
%
\color{red} 
We will denote the empire of $v$ in $\links$ by $\emp v$ (used both as a graph and a set of vertices).
\color{black}

\begin{proposition}[{\cite[Proposition 2.b]{Bellin-vandeWiele-1995}}]
\label{prop:empire propagation}
The empire $\emp v$ is determined by propagation from $v$: 1) through links;\\ \color{red} 2) up towards subformulae;\\ 3) into a tensor if one of its subformulae is in $\emp v-\{v\}$;\\ 4) into a par if all its subformulae are in $\emp v-\{v\}$.%;\\ 5) except never into a par or tensor if $v$ is a subformula.
\end{proposition}

@@

Let a \emph{basic} sequent be one of formulae constructed only over $\1$, $\bot$, and $\tn$.
%
After removing \color{red}dangling \color{black} $\bot$-formulae and replacing subformulae $\1\tn A$ with $A$, basic sequents consist of formulae of the form $1$ or $\bigtn(\bot^n)$ with $n\geq2$.

@@

\aftersection
%
\emph{Non-deterministic constraint logic} \cite{Hearn-Demaine-2005,Demaine-Hearn-2008,GamesPuzzlesAndComputation} is a simple graph-rewriting formalism, used here as a convenient tool for \capsabbrev{pspace}-hardness reduction.
%
A \emph{constraint graph} is a graph with weighted edges and an \emph{inflow constraint}---a natural number that may be taken to be always $2$---on each vertex. A \emph{configuration} is an assignment of directions to the edges of the underlying undirected constraint graph. A rewrite step consists of the reversal of a single edge in a configuration:
rewrite steps must preserve the condition that the total weight of the incoming edges at each vertex is at least its inflow constraint.

\color{red}
Figure~\ref{fig:NCL example} shows an example rewrite sequence in a constraint graph.
%
The central node has inflow constraint 2, the thick blue (dark grey) edge has weight 2, and the thinner red (light grey) edges have weight 1.
\color{black}

\begin{figure}\color{red}
$
\quad
\vc{\begin{tikzpicture}[x=3mm,y=3mm]
	\draw[ncl2,->-] (0,0) -- (3,0);
	\draw[ncl1,->-] (3,0) -- (5,2);
	\draw[ncl1,->-] (3,0) -- (5,-2);
	\node[nclnode] at (3,0) {};
\end{tikzpicture}}
\,\perm*\hfill
\vc{\begin{tikzpicture}[x=3mm,y=3mm]
	\draw[ncl2,->-] (0,0) -- (3,0);
	\draw[ncl1,-<-] (3,0) -- (5,2);
	\draw[ncl1,->-] (3,0) -- (5,-2);
	\node[nclnode] at (3,0) {};
\end{tikzpicture}}
\,\perm*\hfill
\vc{\begin{tikzpicture}[x=3mm,y=3mm]
	\draw[ncl2,->-] (0,0) -- (3,0);
	\draw[ncl1,-<-] (3,0) -- (5,2);
	\draw[ncl1,-<-] (3,0) -- (5,-2);
	\node[nclnode] at (3,0) {};
\end{tikzpicture}}
\,\perm*\hfill
\vc{\begin{tikzpicture}[x=3mm,y=3mm]
	\draw[ncl2,-<-] (0,0) -- (3,0);
	\draw[ncl1,-<-] (3,0) -- (5,2);
	\draw[ncl1,-<-] (3,0) -- (5,-2);
	\node[nclnode] at (3,0) {};
\end{tikzpicture}}
\quad
$

\smallskip
\caption{A series of reconfiguration steps in a constraint graph}
\label{fig:NCL example}
\end{figure}

@@

\begin{definition}
\label{def:standard linkings}
For $W=W_n(i,j)=A\parr B\parr C$ a weight element,
\begin{enumerate}
	\item
for a constraint element $C=C_n(i)=X\tn Y$, the \emph{standard linking} for the sequent $W,C$ links the first $\bot$ in $A$ to the first $\1$ in $X$, the first $\bot$ in $B$ and $C$ each to the first $\1$ in $Y$, and each remaining $\bot$ in $A,B,C$ to a remaining $\1$ in $X,Y$ in their order of occurrence;
	\item
for $C=C_n(j)=X\tn Y$, the \emph{standard linking} for $W,C$ is defined as above, except the first $\bot$ in $B$ links to the first $\1$ in $X$;
	\item
the \emph{standard linking} for the sequent $W,\1^{3n+4}$ links the first $\bot$ in $A$, $B$, and $C$ to the first $\1$, and each remaining $\bot$ to a remaining $\1$ in order of occurrence.
\end{enumerate}
\end{definition}
